---
title: "Analyze_trip_data"
author: "Juliette"
date: "12/03/2021"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Analyzing cyclistic fictional trip data

This documents explores how different customers for a fictional company called Cyclistic are using their ride-share program in the last twelve months from August 2020 to July 2021. The data was stored in individual excel files by month. The data is owned and licensed by Motivate International Inc (https://divvy-tripdata.s3.amazonaws.com/index.html)


## Load packages

Here I load the packages for analysis

```{r packages}
library(tidyverse)
library(dplyr)
library(ggplot2)
library(lubridate)
library(scales)
```

### Load and merge files

Here I load and merge the 12 excel files

```{r load and merge 12 excel files}
library(data.table)
setwd("C:/Users/julie/OneDrive/Documents/trip/Cyclist")
files <- list.files(pattern = ".csv")
data <- lapply(files, fread, sep= ",")
trip_data <- rbindlist(data)
```

### Clean data

Here I start cleaning the data, removing empty columns, spaces, looking for duplicates

```{r clean data}
### remove N/A columns and empty columns
trip_data2 <- filter(trip_data, start_station_id != "NA" & end_station_id!="NA")
trip_data3 <-  filter(trip_data, start_station_name != "" & end_station_name!="")

### check for  duplicates
ride_id <- trip_data3 %>%
  count(ride_id) %>%
  filter(n > 1) # no duplicates found
```

## Transform data

Here I start transforming the data for next step in analysis

### Find time difference in sec between ended_at and started_at columns

Here I find ride length between start and end stations in minutes

```{r}
trip_data3$ride_length <- difftime(
  trip_data3$ended_at, 
  trip_data3$started_at,
  units = "mins"
) 
```

### Remove ride length < 0

Here I remove values less than zero that can negatively skew the results

```{r}
trip_data3 <- trip_data3 %>%
  filter(!(ride_length < 0))
```

### Create column for day of the week, month, year

Here I create three columns, one for the days of the week, month and year

```{r}
trip_data3$day_of_week <- format(trip_data3$started_at,
                                "%A")
trip_data3$day_of_week <- ordered(trip_data3$day_of_week, levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))
trip_data3$month <- format(trip_data3$started_at, "%m")
trip_data3$year <- format(trip_data3$started_at, "%y")
trip_data3$hour <- format(trip_data3$started_at, "%H:%M:%S")

```

### Select variables to work with

Here I select subset of variables to work with in next step of analysis

```{r}
trip_variables <- trip_data3 %>%
  select(ride_id, rideable_type,started_at, start_station_name, end_station_name, member_casual,hour, ride_length, day_of_week, year, month)
```


## Analyze members and casual riders

Here I start looking for similarities and differences among member and casual riders

### Common start station trips for members

Here I create a data frame that includes common start station trips for members

```{r start station trips for members}
trip_variables %>%
  select (start_station_name, member_casual) %>%
  group_by(start_station_name)%>%
  filter(member_casual == "member")%>%
  mutate(trips = n()) %>%
  distinct(start_station_name, .keep_all = TRUE)
```


### Common end station trips for members

Here I create a data frame for common end station trips for members

```{r common end station trips for members}
 trip_variables %>%
  select (end_station_name, member_casual) %>%
  group_by(end_station_name)%>%
  filter(member_casual == "member")%>%
  mutate(trips = n()) %>%
  distinct(end_station_name, .keep_all = TRUE)
```

### Common start station trips for casual riders

Here I create a data frame for common start station trips among casual riders

```{r common start station trips for casual riders}
 trip_variables %>%
  select (start_station_name, member_casual)%>%
  group_by (start_station_name)%>%
  filter(member_casual == "casual")%>%
  mutate(trips = n()) %>%
  distinct(start_station_name, .keep_all= TRUE)
```

### Common end station trips for casual riders

Here I create a data frame for common end stations trips for casual riders

```{r common end station trips for casual riders}
trip_variables %>%
  select (end_station_name, member_casual)%>%
  group_by (end_station_name)%>%
  filter(member_casual == "casual")%>%
  mutate(trips = n()) %>%
  distinct(end_station_name, .keep_all= TRUE)
```


## Summarize number of trips by month

Here I create a data frame that summarizes number of trips by month

```{r}
trip_variables %>%
  select(
    member_casual, 
    month) %>%
  group_by(member_casual,) %>%
  mutate( trips = n()) %>%
  distinct(
    month, 
    member_casual,
    .keep_all = TRUE
  )
```


## Dataframe: summarize number of trips by year

Here I create a data frame that summarizes number of trips by year

```{r}
 trip_variables %>%
  select(
    member_casual, year) %>%
  group_by(member_casual) %>%
  mutate(trips = n()) %>%
  distinct(
    year, 
    member_casual,
    .keep_all = TRUE
  )
```

## Visualization

### Summarize member casual ride length, group by member_casual

Here I summarize ride length by customer type

```{r}
trip_variables %>%
  select(member_casual, ride_length)%>%
  group_by(member_casual) %>%
  summarize(mean(ride_length), sd(ride_length)) # casual members ride longer than members

```

### Visualize member and casual riders

Here I visualize member and casual riders

```{r}
ggplot(data = trip_variables) +
  geom_bar(mapping = aes(x = member_casual))+
labs(title = "Trips by Customer Type" ) +
  xlab("Customer") + ylab("Number of trips") +
  theme(text = element_text(size=15), axis.text = element_text(size=15),legend.text=element_text(size=15))
```


### Group by rideable_type

Here I group customer by type of bikes used

```{r}
trip_variables %>%
  group_by(member_casual, rideable_type) %>%
  summarize(mean(ride_length))

```

### Types of bikes riders use

Here I plot types of bikes used by riders

```{r}
ggplot(data = trip_variables) +
  geom_bar(mapping = aes(x = rideable_type))+
labs(title = " Types of Bikes" ) +
  xlab("Bike Type") + ylab("Number of trips") +
  theme(text = element_text(size=15), axis.text = element_text(size=15),legend.text=element_text(size=15)) 
# most popular bike is the classic bike followed, by the docked bike, and lastly the electric bike
```

### Compare  member and casual riders to types of bikes used

Here I compare member and casual riders to types of bikes used

```{r}
ggplot(data = trip_variables) + 
  geom_bar(mapping = aes(x = member_casual, fill=rideable_type))+
labs(title = "Bike Usage by Customers" ) +
  xlab("Customer") + ylab("Number of trips") +
  theme(text = element_text(size=15), axis.text = element_text(size=15),legend.text=element_text(size=15))
# members take more trips with the classic bike and docked bike compared to casual customers
```

### Plot  day of week

Here I plot the week days to find trends

```{r}
ggplot(data = trip_variables
       ) +
  geom_bar(mapping = aes(x = day_of_week))+ # saturday and sunday has most riders
labs(title = "Trips by Week Day" )+
  xlab("Week Day") + ylab("Number of trips")+
  theme(text = element_text(size=15), axis.text = element_text(size=10),legend.text=element_text(size=15))
```

### Average ride length per day based on customer type

Here I compare the average ride length per day based on customer type

```{r}
all_trips_v2 <- trip_variables %>% 
  select(day_of_week, member_casual, ride_length) %>%  
  group_by(member_casual, day_of_week, ride_length) %>%  
  summarise(trips = n())%>%
  arrange(member_casual, day_of_week)

all_trips_v2 %>% 
select(day_of_week, member_casual, ride_length)%>%
  group_by(member_casual,day_of_week)%>% 
  summarise(trips = n()
            ,average_duration = mean(ride_length))%>% 
  arrange(member_casual, day_of_week)%>% 
  ggplot(aes(x =day_of_week, y = trips, fill = member_casual))+
  geom_col(position = "dodge") +
  labs(title = "Average Ride Length per Day for Customers" , fill=" customer")+
  xlab("Day") + ylab("number of rides")+
  theme(text = element_text(size=15), axis.text = element_text(size=10),legend.text=element_text(size=10)) 
# Average ride length per day is higher for casual riders compared to members
```

### Plot Month

Here I plot month to find trends

```{r}
ggplot(data = trip_variables) +
  geom_bar(mapping = aes(x = month))+ # summer months most popular time of year to ride
labs(title = "Trips by Month" ) +
  xlab("Month") + ylab("Number of trips")
```

### Average Trip Duration for August 2020-July 2021

Here I find the average ride length for customers in from August 2020-July 2021

```{r}
ride_length_stat_2020_2021 <- trip_variables %>% 
    select(year, month,day_of_week, member_casual, ride_length) %>% 
    group_by(year, month,day_of_week, member_casual) %>% 
    summarise(trips = n()
              ,average_duration = round( mean(ride_length) ,digits=1)
              ,sum_duration = round( sum(ride_length), digits =1))%>%
  arrange(year, month, day_of_week, member_casual)
  
  
  ride_length_stat_2020_2021 %>% 
   mutate(year_month = as.Date(paste(year,"-",month,"-1",sep="")) ) %>% 
    group_by(member_casual, year_month) %>% 
    summarise(sum_of_rides = sum(trips )
              ,average_duration = sum(sum_duration) / sum(trips) ) %>% 
    arrange(member_casual, year_month)  %>% 
    ggplot(aes(x = year_month, y = average_duration, group = member_casual, color = member_casual)) +
    geom_line( position = position_dodge(width = 0.9) ) +
    geom_point() +
    labs(title = " Average Trip Duration: August 2020-July 2021" ) +
    xlab("Month") + ylab("Average Duration(mins)") +
    scale_x_date( labels = date_format("%m"), breaks = "1 month", minor_breaks ="1 month") +  
    theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position="bottom", text = element_text(size=12), axis.text = element_text(size=10),legend.text=element_text(size=12))
  # Average ride duration peaked in August and February for both casual and member riders
```

## Start station average trip

Here I analyze station trips

### Group data by start station

Here I first create a data frame that includes start station and number of trips

```{r group data by start station}
trip <- trip_variables%>%
  group_by(start_station_name)%>%
  mutate(trips = n())
```

### Assign variable for mean of trips

Here I assign a variable to find mean of trips

```{r assign variable for mean of trips}
x <- trip %>% 
   summarise( trp = mean(trips))
```

### Select top 10 start stations

Here I assign a variable to find top 10 start stations

```{r select top 10 start stations}
 y <- head( x, 10)

```

### Create horizontal bar chart for top 10 start stations with average trip

Here I create a horizontal bar chart for the top 10 start stations

```{r}
options(repr.plot.width=8, repr.plot.height=3)
 ggplot(y, aes(x = reorder(start_station_name, trp), start_station_name, y = trp)) +
   geom_bar(stat = "identity") +
   coord_flip() + scale_y_continuous(name="Average Trip") +
   scale_x_discrete(name="Start Station")+
   labs(title = "Top 10 start stations average trip")+
   
    theme(axis.text.x = element_text(face="bold", color="black",
                                    size=8, angle=0),
         axis.text.y = element_text(face="bold", color="black",
                                    size=8, angle=0))
```


## Summary of Findings

* Saturday and Sunday has most riders compared to other days of the week
* Average ride length per day is higher for casual riders compared to members
* Average ride length from August 2020 to July 2021 was higher for casual riders compared to members
* Summer months most popular time of year to ride
* Average ride duration peaked in August and February for both casual and member customers
* The most popular bike is the classic bike,  followed by the docked bike and last is the electric bike for all customer types
*  Members take more trips with the classic bike and docked bike compared to casual customers

### Recommendations
* Members take more trips but the average ride length is less compared to casual riders and vice versa. It would be best to target casual riders who take more trips during the week to convert them to members
* Best to target casual customers year around, but especially during summer months
* Average ride duration peaked in August  and February for all customer types, best to target casual riders to convert to member status during these months


